Covariate analysis of ROSMAP Reprocessed RNASeq (with SVA)

### 2016-05-28 code/Rmd/ROSMAP_RNASeq_SVA_Reprocessed.Rmd
Added markdown to process reprocessed RNASeq of ROSMAP data with SVA

---
title: "Covariate analysis of ROSMAP Reprocessed RNASeq"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
```

@@

@@

thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "cogdx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

>>>>

---
title: "Covariate analysis of ROSMAP Reprocessed RNASeq (with SVA)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_SVA_Reprocessed.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx with SVA)')
```

@@

library(sva)

# library(doParallel)
# library(foreach)
# 
# cl = makeCluster(14)
# registerDoParallel(cl)

@@

thisFileName <- 'ROSMAP_RNASeq_SVA_Reprocessed.Rmd'

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = DM1$design[,-grep('cogdx',colnames(DM1$design))] # Get null model by removing variable of interest
MODEL1 = DM1$design

## Calculate the number of surrogate variables
# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]

# Permute and calculate the permuted singular values
PERM.SING.VALS <- foreach (i = 1:100,
         .combine = rbind,
         .packages = c('limma','CovariateAnalysis'),
         .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
  PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
  # Estimate voom weights
  VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)

  PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
  PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
}
num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
NUM.SV = min(num.sv1, num.sv2)

SURR.VAR = sva(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 20)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

# Estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix, design=cbind(MODEL1, SURR.VAR), plot=F)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,VOOM.ADJUSTED.GENE_EXPRESSION$E)
```

### Residual calculation
Calculate weighted residuals and add back "cogdx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("cogdx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
